\thetitle

### 2013-09-29 tfpie/response.txt

RESPONSE: We now make this explicit.

Page 12

Can you quantify the recruitment/retention impact on female students?
Give the reader a sense of what you mean. It must be significant given
that you mention it!

RESPONSE: A quantitative evaluation would be inappropriate in this
situation. We have clarified the point.

----------------------- REVIEW 2 ---------------------

### 2013-09-29 tfpie/tfpie.tex
review 1

An \emph{object} is a value that is a member of this class of data,
which can be constructed with the {\tt new} keyword, a class name, and
the appropriate number of arguments for the fields of the object:
\begin{verbatim}
    (new posn 3 4)
\end{verbatim}
An object understands some set of \emph{messages}.  Simple
structure-like objects understand messages for accessing their fields
and message are sent by using the {\tt send} keyword, followed by an object,
a message name, and some number of arguments:
\begin{verbatim}
    (send (new posn 3 4) x) ;=> 3
    (send (new posn 3 4) y) ;=> 4
\end{verbatim}
The {\tt send} notation is simple, but syntactically heavy.  Once
students are comfortable with {\tt send}, we introduce 
shorthand to make it more convenient, writing {\tt (x . m)} for {\tt
  (send x m)}.  The \emph{dot notation} can be nested, so {\tt (x . m
  . n)} is shorthand for {\tt (send (send x m) n)}.  (The approach of
introducing a simple, uniform syntax and later introducing a
convenient shorthand that would have been confusing to start with
follows the approach of first introducing {\tt cons} and then later {\tt list}
and {\tt quote} in the first semester.)

@@

The structure of this data definition is analogous to the approach of
the previous semester but this example brings to light an important
difference with the functional approach.  The method for computing the
sum of a leaf is defined in the {\tt leaf} class, while the
method for computing the sum of a node is in the {\tt node} class.
When a tree object is sent the {\tt sum} method, there is no function
with a conditional to determine whether the object is a
leaf---instead, the object itself takes care of computing the sum
based on its own {\tt sum} method.  This shift in perspective
is at the core of object-orientation: objects contain their own
behavior and the case analysis previously done in functions is
eliminated.

@@

The move to object-oriented style immediately and naturally leads to
designs that are enabled by organizing programs around objects.  For
example, the state pattern~\cite{samth:GOF} becomes useful almost
immediately.  The programs in figure~\ref{fig:world} animate a rocket
(rendered as a circle in this example) taking off.  An illustrative
follow-up exercise is to animate a rocket that \emph{lands}.  The
natural design is to have two variants for states of the rocket: one
for descending rockets, one for landed rockets (an example is given in
appendix~\ref{sec:appendix}).  While in the functional approach it
is easy to use the state-pattern for the \emph{data} representing a
rocket, it is more difficult to have states of \emph{behavior}.  The
typical solution adds conditionals to all of the event handlers.  In
the object-oriented approach, states of behavior are just as natural
as data.  It is therefore straightforward to design programs with easy
to observe invariants such as ``a landed rocket never changes
position.''  In the functional approach, even such simple properties
are more involved to establish.

@@

The problems of teaching Java in introductory courses have been
well-explored; we mention only a few related directions here.
DrJava~\cite{dvanhorn:Allen2002DrJava} and
BlueJ~\cite{dvanhorn:Hsia2005Taming,dvanhorn:Kolling2003} are
introductory environments for Java, which alleviate some but not all
of the drawbacks we have outlined.
%
Several teaching-oriented graphics libraries for Java have been
proposed~\cite{dvanhorn:Bruce2001Library,dvanhorn:Alphonce2003Using},
but these are significantly more complex than the graphics and
interaction libraries we are able to use in the introductory language
we present.

@@

We have now completed the third iteration of this course, teaching
approximately 35 students each time.  Our experience has been
uniformly positive, and the students have gone on to significant
success in the subsequent courses, despite the curriculum differing
from what the bulk of Northeastern University computer science majors
take.  The class has also had notable success in the recruitment and
retention of female students, as compared to the other versions of the
second-semester course.

>>>>

An \emph{object} is a value that is a member of this class of data,
which can be constructed with the {\tt new} keyword, a class name, and
the appropriate number of arguments for the fields of the object:
\begin{verbatim}
    (new posn 3 4)
\end{verbatim}
An object understands some set of \emph{messages}.  Simple
structure-like objects understand messages for accessing their fields
and messages are sent by using the {\tt send} keyword, followed by an object,
a message name, and some number of arguments:
\begin{verbatim}
    (send (new posn 3 4) x) ;=> 3
    (send (new posn 3 4) y) ;=> 4
\end{verbatim}
The {\tt send} notation is simple, but syntactically heavy.  Once
students are comfortable with {\tt send}, we introduce 
shorthand to make it more convenient, writing {\tt (x . m)} for {\tt
  (send x m)}.  The \emph{dot notation} can be nested, so {\tt (x . m
  . n)} is shorthand for {\tt (send (send x m) n)}.  (The approach of
introducing a simple, uniform syntax and later introducing a
convenient shorthand that would have been confusing to start with
follows the approach of first introducing {\tt cons} and then later {\tt list}
and {\tt quote} in the first semester.)

@@

The structure of this program is analogous to the approach of
the previous semester but this example brings to light an important
difference with the functional approach.  The method for computing the
sum of a leaf is defined in the {\tt leaf} class, while the
method for computing the sum of a node is in the {\tt node} class.
When a tree object is sent the {\tt sum} method, there is no function
with a conditional to determine whether the object is a
leaf---instead, the object itself takes care of computing the sum
based on its own {\tt sum} method.  This shift in perspective
is at the core of object-orientation: objects contain their own
behavior and the case analysis previously done in functions is
eliminated.

@@

The move to object-oriented style immediately and naturally leads to
designs that are enabled by organizing programs around objects.  For
example, the state pattern~\cite{samth:GOF} becomes useful almost
immediately.  The programs in figure~\ref{fig:world} animate a rocket
(rendered as a circle in this example) taking off.  An illustrative
follow-up exercise is to animate a rocket that \emph{lands}.  The
natural design is to have two variants for states of the rocket: one
for descending rockets and one for landed rockets (an example is given in
appendix~\ref{sec:appendix}).  While in the functional approach it
is easy to use the state-pattern for the \emph{data} representing a
rocket, it is more difficult to have states of \emph{behavior}.  The
typical solution adds conditionals to all of the event handlers.  In
the object-oriented approach, states of behavior are just as natural
as data.  It is therefore straightforward to design programs with easy
to observe invariants such as ``a landed rocket never changes
position.''  In the functional approach, even such simple properties
are more involved to establish.

@@

The problems of teaching Java in introductory courses have been
well-explored; we mention only a few related directions here.
DrJava~\cite{dvanhorn:Allen2002DrJava} and
BlueJ~\cite{dvanhorn:Hsia2005Taming,dvanhorn:Kolling2003} are
introductory environments for Java, which alleviate some but not all
of the drawbacks we have outlined.
%
For example, both of these systems immediately present students with
(1) type systems and (2) Java syntax, and (3) do not support the image
values and exact numeric values that we rely on in our course.

Several teaching-oriented graphics libraries for Java have been
proposed~\cite{dvanhorn:Bruce2001Library,dvanhorn:Alphonce2003Using},
but these are significantly more complex than the graphics and
interaction libraries we are able to use in the introductory language
we present.

@@

We have now completed the third iteration of this course, teaching
approximately 35 students each time.  Our experience has been
uniformly positive, and the students have gone on to significant
success in the subsequent courses, despite the curriculum differing
from what the bulk of Northeastern University computer science majors
take.  Anectodally, the class has also had notable success in the
recruitment and retention of female students, as compared to the other
versions of the second-semester course. However, the classes are
sufficiently different as to make a precise comparison impossible.
